---
title: "TeaTime4schools: Joint analysis - bacteria"
subtitle: "07 Beta diversity analysis"
author: "Roey Angel"
email: "roey.angel@bc.cas.cz"
date: "`r Sys.Date()`"
bibliography: references.bib
link-citations: yes
csl: fems-microbiology-ecology.csl
always_allow_html: true
output:
  rmarkdown::github_document:
    toc: true
    toc_depth: 5
    number_sections: false
    dev: "png"
    df_print: "kable"
    keep_html: true
---

```{r libraries, include=F}
# Load libraries
#.libPaths(c('~/R/library', .libPaths())) # Uncomment if you have no write access to R path
library(ragg) # Graphic Devices Based on AGG, CRAN v1.1.2 
library(extrafont)
library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
library(rmarkdown) # Dynamic Documents for R
library(tidyverse) # for dplyr forcats ggplot2 readr tibble
library(broom) # Convert Statistical Analysis Objects into Tidy Data Frames (should be part of tidyverse)
library(gridExtra) # Miscellaneous Functions for "Grid" Graphics
library(cowplot) # wrappers for ggplot
library(ggrepel) # Repulsive Text and Label Geoms for 'ggplot2'
library(magrittr) # pipes
library(scales) # Generic plot scaling methods
library(svglite) # for svg files
library(vegan) # community ecology methods
library(doParallel) # parallel backend for the foreach/%dopar% function
library(vsn) # Variance stabilization and calibration for microarray
library(ade4) # Analysis of Ecological Data
library(ggthemes) #Extra Themes, Scales and Geoms for 'ggplot2'
library(ggsci) #Scientific Journal and Sci-Fi Themed Color 
library(phyloseq) # Handling and analysis of high-throughput phylogenetic sequence data
library(ALDEx2) # Analysis Of Differential Abundance Taking Sample Variation Into Account
library(ggpomological) # Pomological plot themes for ggplot2
```

```{r style settings, include=F}
options(width = 90, knitr.table.format = "html") 
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  dev = "ragg_png",
  fig.ext = "png",
  # dev = c("svglite", "ragg_png"),
  # dev.args = list(svglite = list(bg = 'white', fix_text_size = FALSE), ragg_png = list(bg = 'white')),
  dpi = 300,
#  fig.width = 12,
#  fig.height = 8,
  cache.path = "07_Beta_cache/",
  fig.path = "07_Beta_figures/"
)
f_name <- "DejaVu Sans" #sub("\\s//", "", f_name)
f_size <- 14
font_import(pattern = "DejaVuSans\\.", prompt = FALSE)
loadfonts() # registers fonts
theme_set(theme_bw(base_size = f_size, base_family = f_name))
```

```{r functions, include=F}
PlotLibDist <- function(physeq) {
  ggplot(sample_data(physeq),
         aes(x = Replicate, y = Lib.size, fill = Sample.type)) +
    geom_bar(stat = "identity",
             position = "dodge",
             color = "black") +
    scale_y_log10(
      breaks = trans_breaks("log10", function(x)
        10 ^ x),
      labels = trans_format("log10", math_format(10 ^ .x))
    ) +
    background_grid(major = "xy", minor = "none") +
    # scale_fill_locuszoom() +
    facet_grid(Field ~ .)
}

PlotReadHist <- function(OTUmat, b.width = 10) {
  OTUmat %>%
    t() %>%
    as_tibble() %>%
    gather(key = sample, value = abundance) %>%
    ggplot(aes(abundance)) +
    # geom_histogram(binwidth = 1000) +
    geom_freqpoly(binwidth = b.width) +
    scale_y_log10()
}


PairwiseAdonis <- function(x, factors, sim.function = "vegdist", sim.method = "horn", 
                           p.adjust.m = "BH", reduce = NULL) 
{
  # Taken from: https://github.com/pmartinezarbizu/pairwiseAdonis
  co <- combn(unique(as.character(factors)), 2)
  pairs <- c()
  total.DF <- c()
  F.Model <- c()
  R2 <- c()
  p.value <- c()
  for (elem in 1:ncol(co)) {
    
    if (sim.function == "daisy") {
      x1 = daisy(x[factors %in% c(co[1, elem], co[2, elem]), ],
                 metric = sim.method)
    } else {
      x1 = vegdist(x[factors %in% c(co[1, elem], co[2, elem]), ],
                   method = sim.method)
    }
    ad <- adonis2(x1 ~ fct, 
                  data = data.frame(fct = factors[factors %in% c(co[1, elem], 
                                              co[2, elem])]),
                  permutations = 999)
    # ad <- adonis(x1 ~ factors[factors %in% c(co[1, elem], 
    #     co[2, elem])], permutations = 999) # fails in vegan 2.5-7
    # ad1 <- adonis(x[factors %in% c(co[1, elem], co[2, elem]), ] ~ fct,
    #               data = data.frame(fct = factors[factors %in% c(co[1, elem],
    #                                                         co[2, elem])]),
    #               method="horn",
    #               permutations = 999)
    
    pairs <- c(pairs, paste(co[1, elem], "vs", co[2, elem]))
    # total.DF <- c(total.DF, ad$aov.tab["Total", 1]) # fails in vegan 2.5-7
    total.DF <- c(total.DF, ad["Total", "Df"])
    F.Model <- c(F.Model, ad[1, "F"])
    R2 <- c(R2, ad[1, "R2"])
    p.value <- c(p.value, ad[1, "Pr(>F)"])
  }
  p.adjusted <- p.adjust(p.value, method = p.adjust.m)
  sig = c(rep("", length(p.adjusted)))
  sig[p.adjusted <= 0.05] <- "."
  sig[p.adjusted <= 0.01] <- "*"
  sig[p.adjusted <= 0.001] <- "**"
  sig[p.adjusted <= 1e-04] <- "***"
  pairw.res <- data.frame(pairs, total.DF, F.Model, R2, p.value, 
                          p.adjusted, sig)
  if (!is.null(reduce)) {
    pairw.res <- subset(pairw.res, grepl(reduce, pairs))
    pairw.res$p.adjusted <- p.adjust(pairw.res$p.value, method = p.adjust.m)
    sig = c(rep("", length(pairw.res$p.adjusted)))
    sig[pairw.res$p.adjusted <= 0.05] <- "."
    sig[pairw.res$p.adjusted <= 0.01] <- "*"
    sig[pairw.res$p.adjusted <= 0.001] <- "**"
    sig[pairw.res$p.adjusted <= 1e-04] <- "***"
    pairw.res <- data.frame(pairw.res[, 1:5], sig)
  }
  class(pairw.res) <- c("pwadonis", "data.frame")
  return(pairw.res)
}

STAMPR <-
  function(physeq_obj,
           rank = "Phylum",
           vars2test = c("Spill", "Treatment"),
           sig_pairs,
           threshold = 0.005,
           outputfile = "STAMPR_output") {
    # run a STAMP-like analysis: compare relative abundance differences using two-way analysis, then run a post-hoc test and correct for multiple comparisons
    # The one-way analysis is done using the Kruskal-Wallis Rank Sum Test and the two way analysis is run using the Scheirer Ray Hare test. Both are non-parametric versions of ANOVA on a ranked datasets.
    # The post-hoc test is a Mann Whitney test on each significant pair.
    # Effect sizes (eta^2) are taken from TOMCZAK and TOMCZAK 2014 Trends in Sport Sciences
    # TODO: potentially add effect size calculation to the post-hoc pairs test (MW-U) using eta^2 = Z^2/n
    
    if (length(vars2test) != 1 &
        length(vars2test) != 2)
      {stop('This function only suppurts 1 or two independent variables')}
    
    test_expression <-
      as.formula(paste("Abundance", paste(vars2test, collapse = " + "), sep = " ~ "))
    
    physeq_glom <- tax_glom(physeq_obj,
                            rank,
                            NArm = TRUE)
    physeq_glom_rel <-
      transform_sample_counts(physeq_glom, function(x)
        x / sum(x))
    
    # Test only abundant taxa: group dataframe by rank, calculate median rel. abundance and keep only taxa above threshold
    physeq_glom_rel %>%
      psmelt %>%
      group_by(!!sym(rank)) %>%
      summarise(median = median(Abundance)) %>%
      filter(median >= threshold) %>%
      pull(1) %>%
      as.character() ->
      Taxa2test
    
    physeq_glom_rel_abund <-
      prune_taxa(tax_table(physeq_glom_rel)[, rank] %in% Taxa2test, physeq_glom_rel)
    
    
    if (length(vars2test) == 1) {
      taxa_test_results <-
        bind_cols(Taxa = as(tax_table(physeq_glom_rel_abund)[, rank], "vector"),
                  as.data.frame(matrix(
                    NA,
                    nrow = ntaxa(physeq_glom_rel_abund),
                    ncol = length(sig_pairs) + 2 # sig_pairs is taken from pairwise adonis. In addition the factors are tested alone.
                  )))
      colnames(taxa_test_results) <-
        c(rank,
          paste(vars2test, "- P"),
          paste(vars2test, "- EtaSq"),
          sig_pairs)
      
      taxa_test_stats <-
        bind_cols(Taxa = as(tax_table(physeq_glom_rel_abund)[, rank], "vector"),
                  as.data.frame(matrix(
                    NA,
                    nrow = ntaxa(physeq_glom_rel_abund),
                    ncol = (length(sig_pairs) * 5) # sig_pairs is taken from pairwise adonis
                  )))
      colnames(taxa_test_stats) <-
        c(rank, c(rbind(
          t(str_split_fixed(sig_pairs, " vs ", n = 2)), matrix(rep(
            c("Estimate diff.", "low CI", "high CI"), length(sig_pairs)
          ), ncol = length(sig_pairs))
        ))) # this is ugly but it works well, basically c() flattens a matrix
      
      for (phy_id in seq(ntaxa(physeq_glom_rel_abund))) {
        data2test <-
          bind_cols(Abundance = as.numeric(otu_table(physeq_glom_rel_abund)[, phy_id] * 100),
                as(sample_data(physeq_glom_rel_abund), "data.frame"))
        # kruskal.test(Abundance ~ Climate.Source, data = data2test)
        print(tax_table(physeq_glom_rel_abund)[phy_id, rank])
        
        print(mod_summary <-
                kruskal.test(Abundance ~ Sample.type, data = data2test))
        taxa_test_results[phy_id, c(2)] <-
          mod_summary$p.value[1] # p values
        taxa_test_results[phy_id, c(3)] <-
          (mod_summary$statistic - mod_summary$parameter + 2) / (nrow(data2test) - mod_summary$parameter + 1) # EtaSq (effect size)
        
        for (pair in seq(length(sig_pairs))) {
          pairs2test <- unlist(str_split(sig_pairs[pair], " vs "))
          possibleError <- tryCatch(
            wilcox_mod <-
              wilcox.test(
                as.formula(paste("Abundance", vars2test, sep = " ~ ")),
                data = data2test,
                subset = do.call("%in%", list(get(vars2test), pairs2test)),
                conf.int = TRUE,
                exact = FALSE
              ),
            error = function(e)
              e
          ) # AKA Mann Whitney
          if (inherits(possibleError, 'error')) {
            print(possibleError)
            taxa_test_results[phy_id, pair + 3] <- NA
          } else {
            print(wilcox_mod)
            taxa_test_results[phy_id, pair + 3] <-
              wilcox_mod$p.value
            taxa_test_stats[phy_id, (pair - 1) * 5 + 2] <-
              mean(data2test[unlist(data2test[vars2test]) %in% pairs2test[1],]$Abundance)
            taxa_test_stats[phy_id, (pair - 1) * 5 + 3] <-
              mean(data2test[unlist(data2test[vars2test]) %in% pairs2test[2],]$Abundance)
            taxa_test_stats[phy_id, (pair - 1) * 5 + 4] <-
              wilcox_mod$estimate
            taxa_test_stats[phy_id, (pair - 1) * 5 + c(5, 6)] <-
              wilcox_mod$conf.int[c(1, 2)]
          }
        }
      }
      
    } else {
      factors_combination <- paste(vars2test, collapse = ".")
      sample_data(physeq_obj)[, factors_combination] <-
        paste(get_variable(Ps_obj_filt, vars2test[1]),
              get_variable(Ps_obj_filt, vars2test[2]))
      taxa_test_results <-
        bind_cols(Taxa = tax_table(physeq_glom_rel_abund)[, rank],
                  as.data.frame(matrix(
                    NA,
                    nrow = ntaxa(physeq_glom_rel_abund),
                    ncol = length(sig_pairs) + 6 # sig_pairs is taken from pairwise adonis. In addition the factors are tested alone.
                  )))
      colnames(taxa_test_results) <-
        c(
          rank,
          paste(vars2test[1], "- P"),
          paste(vars2test[1], "- EtaSq"),
          paste(vars2test[2], "- P"),
          paste(vars2test[2], "- EtaSq"),
          paste(paste(vars2test, collapse = " X "), "- P"),
          paste(paste(vars2test, collapse = " X "), "- EtaSq"),
          sig_pairs
        )
      
      taxa_test_stats <-
        bind_cols(Taxa = tax_table(physeq_glom_rel_abund)[, rank],
                  as.data.frame(matrix(
                    NA,
                    nrow = ntaxa(physeq_glom_rel_abund),
                    ncol = (length(sig_pairs) * 5) # sig_pairs is taken from pairwise adonis
                  )))
      colnames(taxa_test_stats) <-
        c(rank, c(rbind(
          t(str_split_fixed(sig_pairs, " vs ", n = 2)), matrix(rep(
            c("Estimate diff.", "low CI", "high CI"), length(sig_pairs)
          ), ncol = length(sig_pairs))
        ))) # this is ugly but it works well, basically c() flattens a matrix
      
      for (phy_id in seq(ntaxa(physeq_glom_rel_abund))) {
        data2test <-
          cbind(Abundance = as.numeric(otu_table(physeq_glom_rel_abund)[, phy_id] * 100),
                as(sample_data(physeq_glom_rel_abund), "data.frame"))
        # kruskal.test(Abundance ~ Climate.Source, data = data2test)
        print(tax_table(physeq_glom_rel_abund)[phy_id, rank])
        
        print(mod_summary <-
                scheirerRayHare(Abundance ~ Sample.type, data = data2test))
        taxa_test_results[phy_id, c(2, 4, 6)] <-
          mod_summary$p.value[1:3] # p values
        taxa_test_results[phy_id, c(3, 5, 7)] <-
          mod_summary$`Sum Sq`[1:3] / sum(mod_summary$`Sum Sq`) # EtaSq (effect size - like ANOVA)
        
        for (pair in seq(length(sig_pairs))) {
          pairs2test <- unlist(str_split(sig_pairs[pair], " vs "))
          possibleError <- tryCatch(
            wilcox_mod <-
              wilcox.test(
                as.formula(paste(
                  "Abundance", factors_combination, sep = " ~ "
                )),
                data = data2test,
                subset = do.call("%in%", list(
                  get(factors_combination), pairs2test
                )),
                conf.int = TRUE,
                exact = FALSE
              ),
            error = function(e)
              e
          ) # AKA Mann Whitney
          if (inherits(possibleError, 'error')) {
            print(possibleError)
            taxa_test_results[phy_id, pair + 7] <- NA
          } else {
            print(wilcox_mod)
            taxa_test_results[phy_id, pair + 7] <-
              wilcox_mod$p.value
            taxa_test_stats[phy_id, (pair - 1) * 5 + 2] <-
              mean(data2test[unlist(data2test[factors_combination]) %in% pairs2test[1],]$Abundance)
            taxa_test_stats[phy_id, (pair - 1) * 5 + 3] <-
              mean(data2test[unlist(data2test[factors_combination]) %in% pairs2test[2],]$Abundance)
            taxa_test_stats[phy_id, (pair - 1) * 5 + 4] <-
              wilcox_mod$estimate
            taxa_test_stats[phy_id, (pair - 1) * 5 + c(5, 6)] <-
              wilcox_mod$conf.int[c(1, 2)]
          }
        }
      }
    }
    
    # Correct for FDR for each comparison pair
    for (pair in seq(2, ncol(taxa_test_results))) {
      # print(pair)
      taxa_test_results[, pair] <-
        p.adjust(taxa_test_results[, pair] , method = "BH") # Benjamini, Y., and Yekutieli, D. (2001). The control of the false discovery rate in multiple testing under dependency. Annals of Statistics 29, 1165–1188.
      # qvalue(p = pull(taxa_test_results[, pair]))
    }
    
    write.csv(taxa_test_results, file = paste0(outputfile, "_", rank, "_Pvals.csv"))
    write.csv(taxa_test_stats, file = paste0(outputfile, "_", rank, "_CI.csv"))
    Taxa_tests <- list(taxa_test_results, taxa_test_stats)
    return(Taxa_tests)
  }

plotSTAMPR <- function(STAMPR_output = Taxa_tests_phylum1, pair = "City - Slope", tax_level = "Phylum", f_size = 14){
  require(ggthemes)
  require(ggpomological)
  if (!is.list(STAMPR_output) | length(STAMPR_output) != 2) {print("The function accepts only lists with two elements")}
  
  pair_number <- which(colnames(STAMPR_output[[1]]) == pair) - 3
  pvals_col <- pair_number + 3
  stats_col <- seq(from = 2, to = (ncol(STAMPR_output[[1]]) - 3) * 5, by = 5)[pair_number - 4]
  
  bind_cols(STAMPR_output[[1]][1], 
            STAMPR_output[[1]][pair] ,
            STAMPR_output[[2]][stats_col:(stats_col + 4)]
  ) %>% 
    mutate(Higher = if_else(.[3] > .[4], colnames(.)[3], colnames(.)[4])) %>% 
    gather("Factor", `Mean proportion (%)`, 3:4) %>% 
    mutate_at(c(tax_level, "Higher", "Factor"), ~fct_rev(.)) -> # fct_rev because coord_flip() reverses the order
    STAMPR_df
    tibble(
      min = seq(
        from = 0.5,
        to = max(as.numeric(pull(STAMPR_df, tax_level))),
        by = 1
      ),
      max = seq(
        from = 1.5,
        to = max(as.numeric(pull(STAMPR_df, tax_level))) + 0.5,
        by = 1
      )) %>% 
        add_column(Shade = rep(c(0, 1), length.out = nrow(.))) %>% 
      mutate_at("Shade", ~as.factor(.)) ->
    # %>% slice(rep(1:n(), each = 2))
    #   mutate(col = ifelse(allyrs == TRUE, 1, 0))
    #   
      # mutate_at("min", ~case_when(col == 1 ~ . - 0.3, 
                                       # TRUE ~ .)) %>% 
      # mutate_at("max", ~case_when(col == 1 ~ . - 0.3, 
                                       # TRUE ~ .)) -> 
      shading
  # ggthemr("grape")
  p1 <- ggplot() +
  geom_rect(data = shading,
            aes(xmin = min, xmax = max, ymin = -Inf, ymax = Inf,
                fill = Shade, alpha = 0.1), fill = rep(c("white", "#E9EDED"), length.out = nrow(shading)), show.legend = F) + 
  geom_col(data = STAMPR_df, 
           mapping = aes(!! sym(tax_level), y = `Mean proportion (%)`, fill = Factor), 
           width = 0.8, 
           position = position_dodge(),
           alpha = 2/3) + 
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_discrete(expand = c(0, 0)) +
    coord_flip() + 
    geom_rangeframe(data = STAMPR_df, aes(!! sym(tax_level), y = `Mean proportion (%)`), sides = "b") +
    # scale_fill_manual(values = Gradient.colours[c(6, 5)])  +
    scale_fill_pomological() +
    theme_tufte(base_size = f_size, base_family = "sans") +
    theme(legend.position = "top",
          legend.justification = 'left',
          legend.title = element_blank()) +
    guides(fill = guide_legend(reverse = TRUE))
          # panel.grid.major.x = element_line(colour = "white"),
          # panel.ontop = TRUE)
  p2 <- ggplot() +
    geom_rect(
      data = shading,
      aes(
        xmin = min,
        xmax = max,
        ymin = -Inf,
        ymax = Inf,
        fill = Shade,
        alpha = 0.1
      ),
      fill = rep(c("white", "#E9EDED"), length.out = nrow(shading)),
      show.legend = F
    ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "slategray", size = 1, alpha = 2/3) + 
    geom_errorbar(data = STAMPR_df,
                  aes(
                    !! sym(tax_level),
                    ymin = `low CI`,
                    ymax = `high CI`,
                    colour = Higher),
                  width = 0.3,
                  alpha = 2/3
    ) +
    geom_point(data = STAMPR_df,
               aes(!! sym(tax_level), 
                   y = `Estimate diff.`, 
                   colour = Higher),
               size = 4,
               alpha = 2/3) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_discrete(
      expand = c(0, 0),
      breaks = STAMPR_df[[tax_level]],
      labels = formatC(STAMPR_df[, 2], format = "e", digits = 2),
      position = "top"
    ) +
    geom_rangeframe(data = STAMPR_df, aes(!! sym(tax_level), y = seq(
      min(`low CI`),
      max(`high CI`),
      length.out = nrow(STAMPR_df)
    )), sides = "b") +
    scale_color_pomological() + 
    theme_tufte(base_size = f_size, base_family = "sans") +
    theme(legend.position = "none", 
          plot.title = element_text(margin = margin(10, 0, 18, 0), hjust = 0.5),
          axis.title.y = element_text(margin = margin(t = 0, r = 0, b = 20, l = 0), angle = 180)) + # BUG: no response
    labs(title = "95% confidence intervals", 
         y = "Difference in mean abundance (%)",
         x = "p-value (corrected)") +
    coord_flip()
  # p1 + p2 + plot_layout(widths = c(1, 2)) # doesn't look as good
  plot_grid(p1, p2, rel_widths = c(2, 3)) + theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
  # ggthemr_reset()
}

DropRareSpecies <- function(Ps_obj = Ps_obj_filt_median, prevalence = 0.1) {
  # calculate sequence prevalence   
  prevdf <- apply( 
    X = otu_table(Ps_obj),
    MARGIN = ifelse(taxa_are_rows(Ps_obj), yes = 1, no = 2),
    FUN = function(x) {sum(x > 0)}
    )
  # Add taxonomy and total read counts to this data.frame
  prevdf <- data.frame(
    Prevalence = prevdf,
    TotalAbundance = taxa_sums(Ps_obj),
    tax_table(Ps_obj)
  )
  
  # Define prevalence threshold as a fraction of the total samples
  prevalenceThreshold <- prevalence * nsamples(Ps_obj)
  prevalenceThreshold
  
  # Execute prevalence filter, using `prune_taxa()` function
  prevdf_phylum_filt <-
    subset(prevdf,
           Phylum %in% get_taxa_unique(Ps_obj, "Phylum"))
  keepTaxa <-
    row.names(prevdf_phylum_filt)[(prevdf_phylum_filt$Prevalence >= prevalenceThreshold)]
  Ps_obj_small <- prune_taxa(keepTaxa, Ps_obj)
  sample_data(Ps_obj_small)$Lib.size <-
    rowSums(otu_table(Ps_obj_small))
  print(Ps_obj)
  print(Ps_obj_small)
  return(Ps_obj_small)
}

CalcALDEx <- function(physeq_obj = Ps_obj_filt_subset, vars2test = "Spill.Treatment", rare_phyla = Rare_phyla, sig_level = 0.05, LFC = 0.322, ...) {
  physeq_obj <- filter_taxa(physeq_obj, function(x) sum(x) > 0, TRUE)
  # physeq_obj <- prune_taxa(sig_taxa, physeq_obj) # remove taxa not significant under the full model
  data2test <- t(otu_table(physeq_obj))
  comparison <- as.character(get_variable(physeq_obj, vars2test))
  ALDEx <- aldex.clr(
    data2test,
    comparison,
    mc.samples = 128,
    denom = "iqlr", # iqlr for slight asymmetry in composition
    verbose = TRUE,
    useMC = TRUE
  ) 
  ALDEx_tt <- aldex.ttest(ALDEx, comparison, paired.test = FALSE) # for two conditions
  ALDEx_effect <- aldex.effect(
    ALDEx,
    comparison,
    include.sample.summary = TRUE,
    verbose = TRUE,
    useMC = TRUE
  ) # estimate effect sizes
  ALDEx2plot <- PrepAlDExData(ALDEx_tt, ALDEx_effect, physeq_obj, sig_level, LFC, Taxa_rank, rare_phyla)
  return(ALDEx2plot)
}

PrepAlDExData <- function(ALDEx_tt, ALDEx_effect, physeq_obj = Ps_obj_filt_subset, sig_level, LFC, Taxa_rank, rare_phyla, ...) {
  ALDEx2plot <- data.frame(ALDEx_tt, ALDEx_effect) # merge results
  # group dataframe by OTU, calculate median rel. abundance
  physeq_obj %>%
    transform_sample_counts(., function(x) x / sum(x) * 100) %>% 
    psmelt() %>%
    group_by(OTU) %>%
    # filter(OTU %in% sig_taxa) %>%
    summarise(baseMean = mean(Abundance)) ->
    baseMean
  
  ALDEx2plot$OTU <- rownames(ALDEx2plot)
  ALDEx2plot %<>% left_join(., baseMean, by = "OTU") # add mean abundance to results table
  ALDEx2plot %<>% cbind(., tax_table(physeq_obj)[taxa_names(physeq_obj) %in% ALDEx2plot$OTU, ], stringsAsFactors = FALSE) # add taxonomy
  # change their name to "Rare"
  
  ALDEx2plot[ALDEx2plot$Phylum %in% rare_phyla, ]$Phylum <- 'Rare' # rare_phyla is calcuted for the taxa box plots
  ALDEx2plot$Significance <- factor("Fail", levels = c("Fail", "Pass")) # define significance factor
  ALDEx2plot$Significance[ALDEx2plot$wi.eBH < sig_level &
                            !is.na(ALDEx2plot$wi.eBH) &
                            abs(ALDEx2plot$effect) > LFC] <- "Pass"
  # ALDEx2plot$Significance <- as.factor(sapply(ALDEx2plot$wi.eBH, function(x) if (is.na(x) | x > 0.05) {x <- "Fail"} else {x <- "Pass"}))
  # Rank by taxa abundance
  ALDEx2plot$Phylum %<>%
    factor(., levels = Taxa_rank$Phylum) %>%  # Taxa_rank is calcuted for the taxa box plots
    fct_relevel(., "Rare", after = Inf)
  return(ALDEx2plot)
}

GGPlotALDExTax <- function(ALDEx2plot, OTU_labels = FALSE, Taxa = "Phylum", Y_val = "effect", sig_level = 0.1) {
  pos <- position_jitter(width = 0.3, seed = 1)
  p <-
    ggplot(ALDEx2plot) +
    geom_point(aes_string(
             x = Taxa,
             y = Y_val,
             colour = "Significance",
             size = "baseMean"),
             position = pos, 
             alpha = 2 / 3, 
             stroke = 0) +
    xlab("") +
    ylab(expression(paste("Effect size (lo", g[2], " fold change)"))) +
    # ylab("Fold change") +
    labs(colour = paste("Significance at \n p <", sig_level), size = "Mean count (%)") +
    theme_grey(base_size = 18,  base_family = "sans") +
    theme(axis.text.x = element_text(angle = 45.0, vjust = 1, hjust = 1)) +
    guides(colour = guide_legend(override.aes = list(size = 5))) +
    scale_colour_manual(values = c(ggpomological:::pomological_base[[7]], ggpomological:::pomological_palette[[1]])) +
    scale_size_continuous(range = c(1, 5), breaks = c(1, 2.5, 5, 10))
  
  if (OTU_labels) {
    p <- p + geom_label_repel(
      aes_string(x = Taxa, y = Y_val),
      size = 6,
      label = sub("Seq_([0-9]+)", "\\1", ALDEx2plot[ALDEx2plot$Significance == "Pass", "OTU"]),
      position = pos,
      data = ALDEx2plot[ALDEx2plot$Significance == "Pass", ],
      # nudge_x = 0.4,
      colour = "#4a4a4a",
      label.size = NA, 
      alpha = 0.75, 
      # fontface = 'bold',
      box.padding = 0.80,
      point.padding = 0.5
      
    )
  }
  return(p)
}

GGPlotOTU <- function(physeq_obj = Ps_obj_filt_subset_New_Oil_Control, vars2test = "Spill.Treatment", OTU_name = "Seq_8"){
  # adapted from https://link.springer.com/protocol/10.1007%2F978-1-4939-8728-3_10
  require("phyloseq")
  require("ggplot2")
  require("ggpomological")
  # Get Proportion transform
  otuRA <- transform_sample_counts(physeq_obj, function(x) x/sum(x))
  gtab <- tibble(
    Count = get_sample(physeq_obj, i = OTU_name),
    Proportion = get_sample(otuRA, i = OTU_name),
    SAMPLE_ID = unlist(get_variable(physeq_obj, "Joint.sample.name")),
    TREATMENT = get_variable(physeq_obj, vars2test)
  )
  suppressWarnings({
    mgtab <- gather(
      gtab,
      "Count", "Proportion",
      key = "Type",
      value = "Abundance",
    )
  })
  # detect if pairs are both zero
  mgtab %<>% group_by(SAMPLE_ID) %>% mutate(., BothZero := all(Abundance == 0))
  
  # Create a dummy min-value for display
  mgtab %<>% group_by(Type) %>% mutate(Zero = min(Abundance[(Abundance > 0.0)], na.rm = TRUE)/10)
  mgtab %<>% group_by(SAMPLE_ID) %>% mutate(Abundance = replace(Abundance, Abundance == 0.0, Zero[Abundance == 0.0]))

  
  pointSize <- 3
  p <- ggplot(data = mgtab,
             mapping = aes(
               x = TREATMENT, 
               y = Abundance,
               color = TREATMENT,
               shape = TREATMENT)) + 
    facet_wrap(~Type, scales = "free_y") +
    # Not both zero
    geom_point(
      data = filter(mgtab, BothZero == FALSE),
      size = pointSize, 
      alpha = 0.8) +
    # Both Zero
    geom_point(
      data = filter(mgtab, BothZero == TRUE),
      size = pointSize, 
      alpha = 0.8,
      position = position_jitter(width = 0.2, height = 0)) + 
    geom_path(
      data = filter(mgtab, BothZero == FALSE),
      mapping = aes(group = SAMPLE_ID), 
      color = "darkgray", 
      size = 0.25,
      position = position_jitter(width = 0, height = 0.001)) + 
    geom_text(mapping = aes(label = SAMPLE_ID),
              data = filter(mgtab, TREATMENT == levels(gtab$TREATMENT)[1] & Abundance > Zero),
              color = "black",
              size = 2,
              nudge_x = -0.15) +
    scale_colour_manual(values = c(ggpomological:::pomological_base[[7]], ggpomological:::pomological_palette[[1]])) +
    # scale_y_sqrt() +
    scale_y_log10() +
    theme_bw() + 
    theme(text = element_text(size = f_size),
          legend.position = "none") +
    scale_size_continuous(range = c(2, 5))
    # ggtitle(paste("Abundance plot for OTU", OTU))
  return(p)
}

GGPlotTopOTUs <- function(physeq_obj = Ps_obj_filt_subset_New_Oil_Control, ALDEx_obj = ALDEx2plot_New_Oil_Control, vars2test = "Spill.Treatment", tax_level = "Order", rank_by = "effect", Ntop = 12){
  # adapted from https://link.springer.com/protocol/10.1007%2F978-1-4939-8728-3_10
  require("phyloseq")
  require("ggplot2")
  require("ggpomological")
  
  # Rank OTUs according to rank_by
  ALDEx_obj %>% 
  filter(Significance == "Pass") %>% 
  dplyr::select(OTU, !! sym(tax_level), baseMean, effect) %>%
  arrange(desc(abs(!! sym(rank_by)))) -> OTU_rank
    if (nrow(OTU_rank) < Ntop) Ntop <- nrow(OTU_rank)
  OTU_rank %>% .[1:Ntop, ] %>% pull(OTU) -> OTU_names
  
  # Get rel. abund. transform
  otuRA <- transform_sample_counts(physeq_obj, function(x) x/sum(x) * 100)

  if (is.na(OTU_names[1]) | is.null(OTU_names[1])) {
    message("No significant differentially abundant OTUs to display")
  } else {
    if (length(OTU_names) == 1) {
      gtab <- tibble(
        OTU =  as_factor(OTU_names),
        # Count = gather(as_tibble(get_sample(physeq_obj, i = OTU_names)))$value,
        Rel_abundance = gather(as_tibble(get_sample(otuRA, i = OTU_names)))$value,
        SAMPLE_ID = as_factor(rep(unlist(get_variable(otuRA, "Joint.sample.name")), Ntop)),
        TREATMENT = as_factor(rep(get_variable(otuRA, vars2test), Ntop)))
      } else {
      gtab <- tibble(
        OTU =  as_factor(gather(as_tibble(get_sample(otuRA, i =
                                                       OTU_names)))$key),
        # Count = gather(as_tibble(get_sample(otuRA, i = OTU_names)))$value,
        Rel_abundance = gather(as_tibble(get_sample(otuRA, i = OTU_names)))$value,
        SAMPLE_ID = as_factor(rep(unlist(get_variable(otuRA, "Joint.sample.name")), Ntop)),
        TREATMENT = as_factor(rep(get_variable(otuRA, vars2test), Ntop)))
    }

  gtab %<>% group_by(SAMPLE_ID, OTU) %>% mutate(., BothZero := all(Rel_abundance == 0))
  
  # Create a dummy min-value for display
  gtab %<>% group_by(OTU) %>% mutate(Zero = min(Rel_abundance[(Rel_abundance > 0.0)], na.rm = TRUE)/10)
  gtab %<>% group_by(SAMPLE_ID, OTU) %>% mutate(Rel_abundance = replace(Rel_abundance, Rel_abundance == 0.0, Zero[Rel_abundance == 0.0]))
  gtab$SAMPLE_ID
  
  pointSize <- 3
  p <- ggplot(data = gtab,
             mapping = aes(
               x = TREATMENT, 
               y = Rel_abundance,
               color = TREATMENT,
               shape = TREATMENT)) + 
    facet_wrap(~OTU, scales = "free_y") +
    # Not both zero
    geom_point(
      data = filter(gtab, BothZero == FALSE),
      size = pointSize, 
      alpha = 0.8) +
    # Both Zero
    geom_point(
      data = filter(gtab, BothZero == TRUE),
      size = pointSize, 
      alpha = 0.8,
      position = position_jitter(width = 0.2, height = 0)) + 
    geom_path(
      data = filter(gtab, BothZero == FALSE),
      mapping = aes(group = SAMPLE_ID),
      color = "darkgray",
      size = 0.25,
      position = position_jitter(width = 0, height = 0.001)) +
    geom_text(mapping = aes(label = SAMPLE_ID),
              data = filter(gtab, TREATMENT == levels(gtab$TREATMENT)[1] & Rel_abundance > Zero),
              color = "black",
              size = 2,
              nudge_x = -0.15) +
    scale_colour_manual(values = c(ggpomological:::pomological_base[[7]], ggpomological:::pomological_palette[[1]])) +
    # scale_y_sqrt() +
    scale_y_log10() +
    theme_bw() + 
    theme(text = element_text(size = 14),
          legend.position = "none") +
    scale_size_continuous(range = c(2, 5)) +
    ylab("Abundance (%)")
    # ggtitle(paste("Abundance plot for OTU", OTU))
  return(p)
  }
}

```
[roey.angel@bc.cas.cz](mailto: roey.angel@bc.cas.cz)  

## Beta diversity analysis
This analysis explores the alpha-diversity ditribution patters in the different samples, based on the DADA2-produced sequences. 

### Setting general parameters:
```{r general parameters}
set.seed(1000)
min_lib_size <- 5000
data_path <- "./DADA2_pseudo/"
Ps_file <- "TeaTime4Schools_16S_filt3_wTree.RDS"
Proj_name <- "TeaTime4Schools"
```

### Load phyloseq object
This phyloseq object was created in [06_Phylogentic_analysis.html](06_Phylogentic_analysis.html). 
The Ps_obj_filt object excludes contaminants and all sequences classified as eukaryota, chloroplast, mitochondria or unknown but still includes taxa with low prevalence 
```{r load phyloseq, cache=T}
readRDS(file = paste0(data_path, Ps_file)) %>%
  subset_samples(., Field != "Unburied") %>% # drop unburied samples
  prune_samples(sample_sums(.) > min_lib_size, .) %>% # remove samples  < min_lib_size
  filter_taxa(., function(x) sum(x) > 0, TRUE) -> # drop taxa with 0 abundance
  Ps_obj_filt
```

```{r}
qplot(rowSums(otu_table(Ps_obj_filt)), geom = "histogram") + 
  xlab("Library size")
qplot(log10(rowSums(otu_table(Ps_obj_filt)))) +
  xlab("Logged library size")
```

### Standardize abundances to the median sequencing depth (and convert to proportion)
```{r median, cache=T}
adonis(
  otu_table(Ps_obj_filt) ~ Lib.size,
  data =
    as(sample_data(Ps_obj_filt), "data.frame"),
  method = "horn",
  permutations = 999
)

Ps_obj_filt %>%
  otu_table(.) %>%
  as(., "matrix") %>%
  rowSums() %>% 
  median() ->
  total
standf = function(x, t = total) round(t * (x / sum(x)))
Ps_obj_filt_median <- transform_sample_counts(Ps_obj_filt, standf)  # Standardize abundances to median sequencing depth

Ps_obj_filt_median_rel <- transform_sample_counts(Ps_obj_filt_median, function(x) x / sum(x)) # convert to relative abundance (just incase it's explicitly needed)

sample_data(Ps_obj_filt_median)$Lib.size <- sample_sums(Ps_obj_filt_median)

qplot(rowSums(otu_table(Ps_obj_filt_median)), geom = "histogram") + 
  xlab("Library size")

PlotLibDist(Ps_obj_filt_median)
```

```{r median diag plots, cache=T, fig.width=4, fig.asp=.8}
PlotReadHist(as(otu_table(Ps_obj_filt_median), "matrix"))
notAllZero <- (rowSums(t(otu_table(Ps_obj_filt_median))) > 0)
meanSdPlot(as(log2(t(otu_table(Ps_obj_filt_median))[notAllZero, ] + 1), "matrix"))
```

### Variance partitioning models and ordinations
#### Partitioning the data using discrete distance
```{r var-part all ,cache=T}
adonis(
  otu_table(Ps_obj_filt_median) ~ Lib.size,
  data =
    as(sample_data(Ps_obj_filt_median), "data.frame"),
  method = "horn",
  permutations = 999
)
adonis(
  otu_table(Ps_obj_filt_median_rel) ~ Field + Sample.type * Season,
  data =
    as(sample_data(Ps_obj_filt_median_rel), "data.frame"),
  method = "horn",
  permutations = 999
)
```

```{r pairwise var-part all ,cache=T}
# sample_data(Ps_obj_filt_median)$Type.season <- paste(sample_data(Ps_obj_filt)$Sample.type, sample_data(Ps_obj_filt)$Season) # too many comparisons

# Ps_obj_filt_median_s <- prune_taxa(names(sort(taxa_sums(Ps_obj_filt_median), TRUE)[1:100]), Ps_obj_filt_median) # for testing

# Compare sample-type pairs
mod_pairwise1 <- PairwiseAdonis(
  otu_table(Ps_obj_filt_median),
  sample_data(Ps_obj_filt_median)$Sample.type,
  sim.function = "vegdist",
  sim.method = "horn",
  p.adjust.m = "BH"
)
print(mod_pairwise1)

(sig_pairs1 <- as.character(mod_pairwise1$pairs[mod_pairwise1$p.adjusted < 0.05]))
# (meaningful_sig_pairs1 <- c("Green tea vs Rooibos", "Green tea vs Soil", "Rooibos vs Soil"))

# compare seasons
mod_pairwise2 <- PairwiseAdonis(
  otu_table(Ps_obj_filt_median),
  sample_data(Ps_obj_filt_median)$Season,
  sim.function = "vegdist",
  sim.method = "horn",
  p.adjust.m = "BH"
)
print(mod_pairwise2)

(sig_pairs2 <- as.character(mod_pairwise2$pairs[mod_pairwise2$p.adjusted < 0.05]))
# (meaningful_sig_pairs2 <- c("Green tea vs Rooibos", "Green tea vs Soil", "Rooibos vs Soil"))
```

##### Calculate ordinations
```{r ordinate all, cache=T, fig.height=12, out.height="40%"}
Ps_obj_ord1 <- ordinate(Ps_obj_filt_median, "CAP", "horn", formula = Ps_obj_filt_median ~  Field + Sample.type * Season)
evals <- eigenvals(Ps_obj_ord1) # /sum(eigenvals(Ps_obj_ord)) * 100

Ps_obj_filt_median %>% 
  plot_ordination(., Ps_obj_ord1, type = "samples", shape = "Field", color = "Sample.type", justDF = TRUE) %>% 
  mutate_at(., "Season", ~fct_relevel(., "Winter", "Spring", "Summer", "Autumn")) %>% 
  mutate_at(., "Sample.type", ~fct_relevel(., "Soil", "Green tea", "Rooibos")) %>% 
  dplyr::rename(., `Sample type` = Sample.type) ->
  ord_df

# p_ord.file <- "PCoA_bray"
# svglite(paste0(p_ord.file, ".svg"),
#         width = 10, height = 7.2)

p_ord <- ggplot(ord_df,
             aes(
               x = CAP1,
               y = CAP2,
               shape = Field,
               color = `Sample type`
             )) +
  geom_point(size = 4, alpha = 2 / 3) +
  theme_bw(base_size = 14) +
  scale_colour_manual(values = pal_locuszoom("default")(3)[c(2, 3, 1)]) +
  stat_ellipse(
    aes(x = CAP1, y = CAP2, group = `Sample type`),
    alpha = 0.5,
    type = "norm",
    level = 0.95,
    linetype = 2,
    inherit.aes = FALSE
  ) +
  labs(x = sprintf("CAP1 [%s%%]", round(evals[1], 1)), 
       y = sprintf("CAP2 [%s%%]", round(evals[2], 1))) +
  coord_fixed(sqrt(evals[2] / evals[1])) +
  facet_wrap(~ Season)

# # Now add the environmental variables as arrows
# arrowmat <- scores(Ps_obj_ord1, display = "bp") # bipplot arrows
# # Add labels, make a data.frame
# arrowdf <- data.frame(labels = rownames(arrowmat), arrowmat)
# 
# arrowdf %<>%
#   mutate(labels = fct_recode(labels,
#     "Braunerde" = "FieldBraunerde",
#     "Kolluvisol" = "FieldKolluvisol",
#     "Rooibos" = "Sample.typeRooibos",
#     "Soil" = "Sample.typeSoil"
#   ))
# arrowdf %<>% dplyr::slice(., c(1:4))
# 
# # Define the arrow aesthetic mapping
# arrow_map = aes(xend = CAP1, yend = CAP2, x = 0, y = 0, shape = NULL, color = NULL, 
#     label = labels)
# label_map = aes(x = 1.2 * CAP1, y = 1.2 * CAP2, shape = NULL, color = NULL, 
#     label = labels)
# # Make a new graphic
# arrowhead = arrow(length = unit(0.05, "npc"))
# p_ord <- p_ord +
#   geom_segment(
#     arrow_map,
#     size = 0.5,
#     data = arrowdf,
#     color = "gray",
#     arrow = arrowhead
#   ) + 
#   geom_text(label_map, size = 2, data = arrowdf)
print(p_ord)
# dev.off()

# ggsave(
#   paste0(p_ord.file, ".png"),
#   p_ord,
#   device = "png",
#   width = 10,
#   height = 6
# )
# gz(paste0(p_ord.file, ".svg"), paste0(p_ord.file, ".svgz"))
```

#### Partitioning the data using phylometric distance
```{r var-part all - phylo ,cache=T}
Unifrac_mat <- UniFrac(Ps_obj_filt_median, 
                       weighted = TRUE, 
                       normalized = TRUE, 
                       parallel = TRUE, 
                       fast = TRUE)

adonis(
  Unifrac_mat ~ Lib.size,
  data =
    as(sample_data(Ps_obj_filt_median), "data.frame"),
  permutations = 999
)
adonis(
  Unifrac_mat ~ Field + Sample.type * Season,
  data =
    as(sample_data(Ps_obj_filt_median), "data.frame"),
  permutations = 999
)
```

##### Calculate ordinations
```{r ordinate all - phylo, cache=T, fig.height=12, out.height="40%"}
Ps_obj_ord2 <- ordinate(Ps_obj_filt_median, "PCoA", "unifrac", weighted = TRUE, formula = Ps_obj_filt_median ~ Field + Sample.type * Season)
evals <- Ps_obj_ord2$values$Eigenvalues[1:2] # /sum(eigenvals(Ps_obj_ord)) * 100

Ps_obj_filt_median %>% 
  plot_ordination(., Ps_obj_ord2, type = "samples", shape = "Field", color = "Sample.type", justDF = TRUE) %>% 
  mutate_at(., "Season", ~fct_relevel(., "Winter", "Spring", "Summer", "Autumn")) %>% 
  mutate_at(., "Sample.type", ~fct_relevel(., "Soil", "Green tea", "Rooibos")) %>% 
  dplyr::rename(., `Sample type` = Sample.type) ->
  ord_df2

# p_ord2.file <- "PCoA_bray"
# svglite(paste0(p_ord2.file, ".svg"),
#         width = 10, height = 7.2)

p_ord2 <- ggplot(ord_df2,
             aes(
               x = Axis.1,
               y = Axis.2,
               shape = Field,
               color = `Sample type`
             )) +
  geom_point(size = 4, alpha = 2 / 3) +
  theme_bw(base_size = 14) +
  scale_colour_manual(values = pal_locuszoom("default")(3)[c(2, 3, 1)]) +
  stat_ellipse(
    aes(x = Axis.1, y = Axis.2, group = `Sample type`),
    alpha = 0.5,
    type = "norm",
    level = 0.95,
    linetype = 2,
    inherit.aes = FALSE
  ) +
  labs(x = sprintf("Axis1 [%s%%]", round(evals[1], 1)), 
       y = sprintf("Axis2 [%s%%]", round(evals[2], 1))) +
  coord_fixed(sqrt(evals[2] / evals[1])) +
  facet_wrap( ~ Season)

# plot_grid(p_ord,p_ord2)
print(p_ord2)
# dev.off()

# ggsave(
#   paste0(p_ord2.file, ".png"),
#   p_ord2,
#   device = "png",
#   width = 10,
#   height = 6
# )
# gz(paste0(p_ord2.file, ".svg"), paste0(p_ord2.file, ".svgz"))
```

### Variance partitioning models and ordinations - soils excluded
#### Partitioning the data using discrete distance
```{r var-part soil ,cache=T}
Ps_obj_filt_median_tea <- subset_samples(Ps_obj_filt_median, Sample.type != "Soil")

adonis(
  otu_table(Ps_obj_filt_median_tea) ~ Lib.size,
  data =
    as(sample_data(Ps_obj_filt_median_tea), "data.frame"),
  method = "horn",
  permutations = 999
)
adonis(
  otu_table(Ps_obj_filt_median_tea) ~ Field + Sample.type * Season,
  data =
    as(sample_data(Ps_obj_filt_median_tea), "data.frame"),
  method = "horn",
  permutations = 999
)
```

```{r pairwise var-part teas ,cache=T}
# sample_data(Ps_obj_filt_median)$Type.season <- paste(sample_data(Ps_obj_filt)$Sample.type, sample_data(Ps_obj_filt)$Season) # too many comparisons

# Ps_obj_filt_median_s <- prune_taxa(names(sort(taxa_sums(Ps_obj_filt_median), TRUE)[1:100]), Ps_obj_filt_median) # for testing

# Compare sample-type pairs
mod_pairwise3 <- PairwiseAdonis(
  otu_table(Ps_obj_filt_median_tea),
  sample_data(Ps_obj_filt_median_tea)$Sample.type,
  sim.function = "vegdist",
  sim.method = "horn",
  p.adjust.m = "BH"
)
print(mod_pairwise3)

(sig_pairs3 <- as.character(mod_pairwise3$pairs[mod_pairwise3$p.adjusted < 0.05]))
# (meaningful_sig_pairs3 <- c("Green tea vs Rooibos", "Green tea vs Soil", "Rooibos vs Soil"))

# compare seasons
mod_pairwise4 <- PairwiseAdonis(
  otu_table(Ps_obj_filt_median_tea),
  sample_data(Ps_obj_filt_median_tea)$Season,
  sim.function = "vegdist",
  sim.method = "horn",
  p.adjust.m = "BH"
)
print(mod_pairwise4)

(sig_pairs4 <- as.character(mod_pairwise4$pairs[mod_pairwise4$p.adjusted < 0.05]))
# (meaningful_sig_pairs2 <- c("Green tea vs Rooibos", "Green tea vs Soil", "Rooibos vs Soil"))
```

##### Calculate ordinations
```{r ordinate tea, cache=T, fig.height=12, out.height="40%"}
Ps_obj_ord3 <- ordinate(Ps_obj_filt_median_tea, "CAP", "horn", formula = Ps_obj_filt_median_tea ~  Field + Sample.type * Season)
evals <- eigenvals(Ps_obj_ord3) # /sum(eigenvals(Ps_obj_ord)) * 100

Ps_obj_filt_median_tea %>% 
  plot_ordination(., Ps_obj_ord3, type = "samples", shape = "Field", color = "Sample.type", justDF = TRUE) %>% 
  mutate_at(., "Season", ~fct_relevel(., "Winter", "Spring", "Summer", "Autumn")) %>% 
  mutate_at(., "Sample.type", ~fct_relevel(., "Green tea", "Rooibos")) %>% 
  dplyr::rename(., `Sample type` = Sample.type) ->
  ord_df3

# p_ord.file <- "PCoA_bray"
# svglite(paste0(p_ord.file, ".svg"),
#         width = 10, height = 7.2)

p_ord3 <- ggplot(ord_df3,
             aes(
               x = CAP1,
               y = CAP2,
               shape = Field,
               color = `Sample type`
             )) +
  geom_point(size = 4, alpha = 2 / 3) +
  theme_bw(base_size = 14) +
  scale_colour_manual(values = pal_locuszoom("default")(3)[c(3, 1)]) +
  stat_ellipse(
    aes(x = CAP1, y = CAP2, group = `Sample type`),
    alpha = 0.5,
    type = "norm",
    level = 0.95,
    linetype = 2,
    inherit.aes = FALSE
  ) +
  labs(x = sprintf("Axis1 [%s%%]", round(evals[1], 1)), 
       y = sprintf("Axis2 [%s%%]", round(evals[2], 1))) +
  coord_fixed(sqrt(evals[2] / evals[1])) +
  facet_wrap(~ Season)

print(p_ord3)
# dev.off()

# ggsave(
#   paste0(p_ord.file, ".png"),
#   p_ord,
#   device = "png",
#   width = 10,
#   height = 6
# )
# gz(paste0(p_ord.file, ".svg"), paste0(p_ord.file, ".svgz"))
```

#### Partitioning the data using phylometric distance
```{r var-part tea - phylo ,cache=T}
Unifrac_mat <- UniFrac(Ps_obj_filt_median_tea, 
                       weighted = TRUE, 
                       normalized = TRUE, 
                       parallel = TRUE, 
                       fast = TRUE)

adonis(
  Unifrac_mat ~ Lib.size,
  data =
    as(sample_data(Ps_obj_filt_median_tea), "data.frame"),
  permutations = 999
)
adonis(
  Unifrac_mat ~ Field + Sample.type * Season,
  data =
    as(sample_data(Ps_obj_filt_median_tea), "data.frame"),
  permutations = 999
)
```

##### Calculate ordinations
```{r ordinate tea - phylo, cache=T, fig.height=12, out.height="40%"}
Ps_obj_ord4 <- ordinate(Ps_obj_filt_median_tea, "PCoA", "unifrac", weighted = TRUE, formula = Ps_obj_filt_median_tea ~ Field + Sample.type * Season)
evals <- Ps_obj_ord4$values$Eigenvalues[1:2] # /sum(eigenvals(Ps_obj_ord)) * 100

Ps_obj_filt_median_tea %>% 
  plot_ordination(., Ps_obj_ord4, type = "samples", shape = "Field", color = "Sample.type", justDF = TRUE) %>% 
  mutate_at(., "Season", ~fct_relevel(., "Winter", "Spring", "Summer", "Autumn")) %>% 
  mutate_at(., "Sample.type", ~fct_relevel(., "Green tea", "Rooibos")) %>% 
  dplyr::rename(., `Sample type` = Sample.type) ->
  ord_df4

# p_ord.file <- "PCoA_bray"
# svglite(paste0(p_ord.file, ".svg"),
#         width = 10, height = 7.2)

p_ord4 <- ggplot(ord_df4,
             aes(
               x = Axis.1,
               y = Axis.2,
               shape = Field,
               color = `Sample type`
             )) +
  geom_point(size = 4, alpha = 2 / 3) +
  theme_bw(base_size = 14) +
  scale_colour_manual(values = pal_locuszoom("default")(3)[c(3, 1)]) +
  stat_ellipse(
    aes(x = Axis.1, y = Axis.2, group = `Sample type`),
    alpha = 0.5,
    type = "norm",
    level = 0.95,
    linetype = 2,
    inherit.aes = FALSE
  ) +
  labs(x = sprintf("Axis1 [%s%%]", round(evals[1], 1)), 
       y = sprintf("Axis2 [%s%%]", round(evals[2], 1))) +
  coord_fixed(sqrt(evals[2] / evals[1])) +
  facet_wrap( ~ Season)

# plot_grid(p_ord,p_ord)
print(p_ord4)
# dev.off()

# ggsave(
#   paste0(p_ord.file, ".png"),
#   p_ord,
#   device = "png",
#   width = 10,
#   height = 6
# )
# gz(paste0(p_ord.file, ".svg"), paste0(p_ord.file, ".svgz"))
```

#### Test differences between samples on the phylum level
```{r STAMPR, fig.width=12, fig.height=6, cache=T, results="hide"}
Taxa_tests_phylum1 <- STAMPR(Ps_obj_filt_median, vars2test = "Sample.type", sig_pairs = sig_pairs1, outputfile = paste0(Proj_name, "_Sample.type"))
plotSTAMPR(Taxa_tests_phylum1, pair = "Green tea vs Rooibos")
plotSTAMPR(Taxa_tests_phylum1, pair = "Green tea vs Soil")
plotSTAMPR(Taxa_tests_phylum1, pair = "Rooibos vs Soil")
Taxa_tests_order1 <- STAMPR(Ps_obj_filt_median, vars2test = "Sample.type", rank = "Order", sig_pairs = sig_pairs1, outputfile = paste0(Proj_name, "_Sample.type"))
plotSTAMPR(Taxa_tests_order1, pair = "Green tea vs Rooibos", tax_level = "Order")
plotSTAMPR(Taxa_tests_order1, pair = "Green tea vs Soil", tax_level = "Order")
plotSTAMPR(Taxa_tests_order1, pair = "Rooibos vs Soil", tax_level = "Order")
Taxa_tests_phylum2 <- STAMPR(Ps_obj_filt_median, vars2test = "Season", sig_pairs = sig_pairs2, outputfile = paste0(Proj_name, "_Season"))
plotSTAMPR(Taxa_tests_phylum2, pair = "Winter vs Summer")
Taxa_tests_order2 <- STAMPR(Ps_obj_filt_median, vars2test = "Season", rank = "Order", sig_pairs = sig_pairs2, outputfile = paste0(Proj_name, "_Season"))
plotSTAMPR(Taxa_tests_order2, pair = "Winter vs Summer", tax_level = "Order")
Taxa_tests_phylum4 <- STAMPR(Ps_obj_filt_median_tea, vars2test = "Season", sig_pairs = sig_pairs4, outputfile = paste0(Proj_name, "Teabags_Season"))
plotSTAMPR(Taxa_tests_phylum4, pair = "Winter vs Summer")
Taxa_tests_order4 <- STAMPR(Ps_obj_filt_median_tea, vars2test = "Season", rank = "Order", sig_pairs = sig_pairs4, outputfile = paste0(Proj_name, "Teabags_Season"))
plotSTAMPR(Taxa_tests_order4, pair = "Winter vs Summer", tax_level = "Order")
```

### Differential abundance models
Tag rare phyla (for plotting purposes only)
```{r tag rare phyla, cache=T}
Ps_obj_filt_media_glom <- tax_glom(Ps_obj_filt_median, 
                             "Phylum", 
                             NArm = TRUE) # glomerate to the phylum level
Ps_obj_filt_media_glom_rel <- transform_sample_counts(Ps_obj_filt_media_glom, function(x) x / sum(x)) # transform to rel. ab.
Ps_obj_filt_media_glom_rel_DF <- psmelt(Ps_obj_filt_media_glom_rel) # generate a df
Ps_obj_filt_media_glom_rel_DF$Phylum %<>% as.character() # factor to char

# group dataframe by Phylum, calculate median rel. abundance
Ps_obj_filt_media_glom_rel_DF %>%
  group_by(Phylum) %>%
  summarise(median = median(Abundance)) ->
  medians

# find Phyla whose median rel. abund. is less than 0.5%
Rare_phyla <- medians[medians$median <= 0.005, ]$Phylum

# change their name to "Rare"
Ps_obj_filt_media_glom_rel_DF[Ps_obj_filt_media_glom_rel_DF$Phylum %in% Rare_phyla, ]$Phylum <- 'Rare'

# re-group
Ps_obj_filt_media_glom_rel_DF %>%
  group_by(Phylum) %>%
  summarise(Abundance = sum(Abundance)) %>% 
  arrange(desc(Abundance)) -> Taxa_rank
```


Detect differentially abundant OTUs using ALDEx2 [@fernandes_anova-like_2013]

**Sample type differences**
```{r ALDEx2 - sample.type, cache=T, results = 'asis', fig.width=10, fig.height=6}
significance = 0.05

# run full model 
data2test <- t(otu_table(Ps_obj_filt_median))
# comparison <- as.character(get_variable(Ps_obj_filt_median, "Sample.type"))
#ALDEx_full <- aldex.clr(data2test, comparison, mc.samples = 128, denom = "iqlr", verbose = TRUE, useMC = TRUE) # iqlr for slight assymetry in composition
#ALDEx_full_glm <- aldex.glm(ALDEx_full, comparison, useMC = TRUE) # for more than two conditions
#sig_taxa <- rownames(ALDEx_full_glm)[ALDEx_full_glm$glm.eBH < 0.05] # save names of taxa that are significant under the full model
#write.csv(sig_taxa, file = "Aldex_full_significant_taxa.csv")

# Pairwise comparisons
# 
ALDEx_comparisons <- list()
ALDEx_comparisons$Comparisons <- sig_pairs1

# Ps_obj_filt_median_subset <- prune_taxa(names(sort(taxa_sums(Ps_obj_filt_median), TRUE)[1:100]), Ps_obj_filt_median)

for (j in seq(1, length(ALDEx_comparisons$Comparisons))) {
  # print(j)
  ALDEx_comparisons$Comparisons[j] %>% 
    str_split(., " vs ", simplify = FALSE) %>% 
    unlist() ->
    comparison_string
  Ps_obj_filt_median %>%
    subset_samples(Sample.type == comparison_string[1] | Sample.type == comparison_string[2]) ->
    # subset_samples(Treatment == comparison_string[2] | Treatment == comparison_string[4]) %>%
    # subset_samples(Year == "2016" | Year == "2017" | Year == "2018") ->
    Ps_obj_filt_median_pairwise

#  Remove species with prevalence < 10%
  Ps_obj_filt_median_pairwise_s <- DropRareSpecies(Ps_obj = Ps_obj_filt_median_pairwise, prevalence = 0.1)
    
  sample_data(Ps_obj_filt_median_pairwise_s)$Sample.type %<>% fct_relevel(., str_split(ALDEx_comparisons$Comparisons[j], " vs ", simplify = TRUE)[1])
  
  # make Joint.sample.name for matching OTUs between compared samples (for GGPlotTopOTUs)
  suppressWarnings(
  sample_data(Ps_obj_filt_median_pairwise_s) %<>%
    as(., "data.frame") %>% 
    rownames_to_column() %>% 
    mutate_if(is.factor, as.character) %>% 
    mutate(Joint.sample.name  = paste0(.$Sample.type, "_", .$Field, "_", .$Season, "_", .$Replicate)) %>% 
    mutate_at("Joint.sample.name", ~str_replace_all(., paste0(unique(comparison_string), collapse = "|"), "")) %>% # remove the levels participating in the comparison from the names
    mutate_at("Joint.sample.name", ~str_replace_all(., "^_", "")) %>% # remove first "_"
    column_to_rownames()
  )
    
  ALDEx2plot_pairwise <- CalcALDEx(
    physeq_obj = Ps_obj_filt_median_pairwise_s,
    vars2test = "Sample.type",
    rare_phyla = Rare_phyla,
    sig_level = significance,
    LFC = 0
    )
  ALDEx_comparisons$Results[[j]] <- ALDEx2plot_pairwise # store results
  ALDEx_comparisons$Results[[j]]$Var1 <- str_split(ALDEx_comparisons$Comparisons[j], " vs ", simplify = TRUE)[1]
  ALDEx_comparisons$Results[[j]]$Var2 <- str_split(ALDEx_comparisons$Comparisons[j], " vs ", simplify = TRUE)[2]
  ALDEX_summary <- tibble(Label = c(paste0("⬆", sum( ALDEx_comparisons$Results[[j]]$effect > 0 &  ALDEx_comparisons$Results[[j]]$Significance == "Pass"), " ⬇", sum( ALDEx_comparisons$Results[[j]]$effect < 0 &  ALDEx_comparisons$Results[[j]]$Significance == "Pass"), " (", nrow( ALDEx_comparisons$Results[[j]]), ")")))
  
  ALDEx2plot_pairwise %>%
    filter(Significance == "Pass") %>%
    dplyr::select(OTU, baseMean, effect, Phylum, Class, Order, Family, Genus) %>%
    arrange(desc(abs(effect))) ->
    ALDEx2plot_pairwise_results
    
  # print(ALDEx2plot_pairwise_results %>% 
  #   kable(., digits = c(2), caption = "Significantly different taxa:") %>%
  #   kable_styling(
  #     bootstrap_options = c("striped", "hover", "condensed", "responsive"),
  #     full_width = F
  #   ))
  
  write.csv(ALDEx2plot_pairwise_results, file = paste0("Aldex", "_", paste0(comparison_string, collapse = "_"), ".csv"))
  
  # Plot ALDEX plot
  p1 <- GGPlotALDExTax(ALDEx2plot_pairwise, OTU_labels = FALSE, sig_level = significance) +
    # ggtitle(ALDEx_comparisons$Comparisons[j]) +
    geom_text(
    data    = ALDEX_summary,
    mapping = aes(x = Inf, y = Inf, label = Label),
    hjust   = 1.1,
    vjust   = 1.6
  ) 
  p1 <- p1 + labs(title = ALDEx_comparisons$Comparisons[j])
  print(p1)
  
  # Plot OTU plots
  # GGPlotOTU(Ps_obj_filt_subset_pairwise_s, vars2test = "Spill.Treatment", "Seq_8")
  p2 <- GGPlotTopOTUs(
    physeq_obj = Ps_obj_filt_median_pairwise_s,
    vars2test = "Sample.type",
    ALDEx_obj = ALDEx2plot_pairwise,
    rank_by = "effect",
    Ntop = 12
  )
  print(p2)
}

filter(ALDEx_comparisons$Results[[1]], Significance == "Pass")$OTU[common_taxa <- filter(ALDEx_comparisons$Results[[1]], Significance == "Pass")$OTU %in% filter(ALDEx_comparisons$Results[[2]], Significance == "Pass")$OTU]
```

**Seasonal differences**
Comparing the seasonsal differences in teabag samples only
```{r ALDEx2 - season, cache=T, results = 'asis', fig.width=10, fig.height=6}
significance = 0.05

data2test <- t(otu_table(Ps_obj_filt_median_tea))

# Pairwise comparisons
ALDEx_comparisons <- list()
ALDEx_comparisons$Comparisons <- sig_pairs2

# Ps_obj_filt_median_tea_subset <- prune_taxa(names(sort(taxa_sums(Ps_obj_filt_median_tea), TRUE)[1:100]), Ps_obj_filt_median_tea)

for (j in seq(1, length(ALDEx_comparisons$Comparisons))) {
  # print(j)
  ALDEx_comparisons$Comparisons[j] %>% 
    str_split(., " vs ", simplify = FALSE) %>% 
    unlist() ->
    comparison_string
  Ps_obj_filt_median_tea %>%
    subset_samples(Season == comparison_string[1] | Season == comparison_string[2]) ->
    # subset_samples(Treatment == comparison_string[2] | Treatment == comparison_string[4]) %>%
    # subset_samples(Year == "2016" | Year == "2017" | Year == "2018") ->
    Ps_obj_filt_median_tea_pairwise

#  Remove species with prevalence < 10%
  Ps_obj_filt_median_tea_pairwise_s <- DropRareSpecies(Ps_obj = Ps_obj_filt_median_tea_pairwise, prevalence = 0.1)
    
  sample_data(Ps_obj_filt_median_tea_pairwise_s)$Season %<>% fct_relevel(., str_split(ALDEx_comparisons$Comparisons[j], " vs ", simplify = TRUE)[1])
  
  # make Joint.sample.name for matching OTUs between compared samples (for GGPlotTopOTUs)
  suppressWarnings(
  sample_data(Ps_obj_filt_median_tea_pairwise_s) %<>% 
    as(., "data.frame") %>% 
    rownames_to_column() %>% 
    mutate_if(is.factor, as.character) %>% 
    mutate(Joint.sample.name  = paste0(.$Sample.type, "_", .$Field, "_", .$Season, "_", .$Replicate)) %>% 
    mutate_at("Joint.sample.name", ~str_replace_all(., paste0(unique(comparison_string), collapse = "|"), "")) %>% # remove the levels participating in the comparison from the names
    mutate_at("Joint.sample.name", ~str_replace_all(., "^_", "")) %>% # remove first "_"
    mutate_at("Joint.sample.name", ~str_replace_all(., "__", "_")) %>% # remove double "_"
    column_to_rownames()
  )
    
  ALDEx2plot_pairwise <- CalcALDEx(
    physeq_obj = Ps_obj_filt_median_tea_pairwise_s,
    vars2test = "Season",
    rare_phyla = Rare_phyla,
    sig_level = significance,
    LFC = 0
    )
  ALDEx_comparisons$Results[[j]] <- ALDEx2plot_pairwise # store results
  ALDEx_comparisons$Results[[j]]$Var1 <- str_split(ALDEx_comparisons$Comparisons[j], " vs ", simplify = TRUE)[1]
  ALDEx_comparisons$Results[[j]]$Var2 <- str_split(ALDEx_comparisons$Comparisons[j], " vs ", simplify = TRUE)[2]
  ALDEX_summary <- tibble(Label = c(paste0("⬆", sum( ALDEx_comparisons$Results[[j]]$effect > 0 &  ALDEx_comparisons$Results[[j]]$Significance == "Pass"), " ⬇", sum( ALDEx_comparisons$Results[[j]]$effect < 0 &  ALDEx_comparisons$Results[[j]]$Significance == "Pass"), " (", nrow( ALDEx_comparisons$Results[[j]]), ")")))
  
  ALDEx2plot_pairwise %>%
    filter(Significance == "Pass") %>%
    dplyr::select(OTU, baseMean, effect, Phylum, Class, Order, Family, Genus) %>%
    arrange(desc(abs(effect))) ->
    ALDEx2plot_pairwise_results
    
  # print(ALDEx2plot_pairwise_results %>% 
  #   kable(., digits = c(2), caption = "Significantly different taxa:") %>%
  #   kable_styling(
  #     bootstrap_options = c("striped", "hover", "condensed", "responsive"),
  #     full_width = F
  #   ))
  
  write.csv(ALDEx2plot_pairwise_results, file = paste0("Aldex", "_", paste0(comparison_string, collapse = "_"), ".csv"))
  
  # Plot ALDEX plot
  p1 <- GGPlotALDExTax(ALDEx2plot_pairwise, OTU_labels = FALSE, sig_level = significance) +
    # ggtitle(ALDEx_comparisons$Comparisons[j]) +
    geom_text(
    data    = ALDEX_summary,
    mapping = aes(x = Inf, y = Inf, label = Label),
    hjust   = 1.1,
    vjust   = 1.6
  ) 
  p1 <- p1 + labs(title = ALDEx_comparisons$Comparisons[j])
  print(p1)
  
  # Plot OTU plots
  p2 <- GGPlotTopOTUs(
    physeq_obj = Ps_obj_filt_median_tea_pairwise_s,
    vars2test = "Season",
    ALDEx_obj = ALDEx2plot_pairwise,
    rank_by = "effect",
    Ntop = 12
  )
  print(p2)
}

filter(ALDEx_comparisons$Results[[1]], Significance == "Pass")$OTU[common_taxa <- filter(ALDEx_comparisons$Results[[1]], Significance == "Pass")$OTU %in% filter(ALDEx_comparisons$Results[[2]], Significance == "Pass")$OTU]
```

```{r colophon, eval=T}
sessioninfo::session_info() %>%
  details::details(
    summary = 'Current session info',
    open    = TRUE
 )
```

## References